Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-06-23 manuscript.tex
Add missing details, add abstract

\begin{abstract}

\end{abstract}

\begin{keyword}
  Monte Carlo, depletion
\end{keyword}

@@

In OpenMC, the burnup matrix is constructed based on data inside of a
\emph{depletion chain} file, which includes fundamental data gathered from ENDF
incident neutron, decay, and fission product yield sublibraries. For each
nuclide, this file includes:
\begin{itemize}
  \item What transmutation reactions are possible, their Q values, and their products;
  \item If a nuclide is not stable, what decay modes are possible, their
  branching ratios, and their products; and
  \item If a nuclide is fissionable, the fission products yields at any number
  of incident neutron energies.
\end{itemize}

Depletion chains can be automatically created in the Python API in the following
manner. First, the user supplies a list of ENDF incident neutron, decay, and
fission product yield sublibrary files. The incident neutron files are used to
determine what nuclear reactions result in transmutation and what the
corresponding Q values of the reactions are. The decay files are used to
determine the half-life and possible decay modes of nuclides that are unstable.
The fission product yield sublibraries are used to determine what fission
products are produced when a fission reaction occurs. This information is then
aggregated and written to an XML file that serves as the depletion chain file
that is used at runtime to construct a burnup matrix.

@@

TODO: Discuss missing FPY~\cite{fensin2015pne}.

Using all incident neutron, decay, and fission product yield sublibrary files
from an ENDF-format library, such as ENDF/B-VIII.0, results in a \emph{full}
depletion chain that typically includes thousands of nuclides. Because the use
of a chain with $n$ nuclides will result in an $n\times n$ burnup matrix, there
is an incentive to reduce the number of nuclides in the depletion chain to
minimize both the size of the burnup matrix and the number of nuclides that are
tracked during a transport solve. The depletion chain class in OpenMC's Python
API includes a method for automatically reducing the size of the chain. Starting
with a set of nuclides that must be included, the method will recursively go
through all decay and reaction modes to determine what products may be produced.
Using this method on the ENDF/B-VII.1 library starting from a list of all stable
or very long lived (half-life greater than $10^{15}$ sec) nuclides reduces the
total number of nuclides in the depletion chain from 3,820 to 1,693.

In addition to the full and reduced depletion chains, a simplified depletion
chain based on the VERA depletion benchmark specification~\citep{kim2015casl}
was created and made available for users at \url{https://openmc.org}. This chain
includes 223 nuclides and has adjusted decay branching ratios for some nuclides
to preserve the accuracy of nuclide production rates. \Cref{fig:chains} shows
which nuclides are included in each of the depletion chains.
\begin{figure}
  \centering
  \includegraphics[width=4in]{figures/chains.pdf}
  \caption{Table of nuclides showing which nuclides are included in each of the
    depletion chains based on ENDF/B-VII.1.}
  \label{fig:chains}
\end{figure}

@@

To validate the depletion solver in OpenMC, we have carried out simulations
using the \texttt{openmc.deplete} module as well as Serpent 2.1.31 on two simple
problems. The first problem is a pressurized water reactor (PWR) pincell and the
second problem is a single assembly of a sodium-cooled fast reactor (SFR). These
problems were selected to cover both thermal and fast neutron spectra.

@@

Six coupled transport-depletion simulations were carried out with OpenMC to test
a variety of combinations of depletion chains and methods for handling FPY
energy dependence:
\begin{enumerate}
  \item Full depletion chain, thermal FPYs
  \item Full depletion chain, interpolated FPYs
  \item Reduced depletion chain, thermal FPYs
  \item CASL depletion chain, thermal FPYs
  \item CASL depletion chain, FPYs based on cutoff method
  \item CASL depletion chain, interpolated FPYs
\end{enumerate}
Because Serpent doesn't have an option for changing the depletion chain or the
method for handling FPY energy dependence, a single coupled transport-depletion
simulation with Serpent was carried out. For all simulations, 20 inactive
batches, 200 active batches, and \num{4.0e6} particles per batch were used,
which resulted in an uncertainty on the effective multiplication factor of about
3--4 pcm.

>>>>

\begin{abstract}
  A depletion solver has been implemented in OpenMC and is described herein. The
  depletion solver is implemented in Python and interfaces with OpenMC's
  transport solver through a C++ API, which enables an in-memory
  depletion--transport coupling. Multiple integration methods for advancing in
  time have been implemented and exhibit tradeoffs in cost, accuracy, and memory
  use. For all time integration methods, evaluation of the matrix exponential is
  performed using the incomplete partial fraction form of the Chebyshev rational
  approximation method.

  Simulations of a PWR pincell and an SFR assembly were carried out with OpenMC
  and Serpent. For both problems, the use of a high-fidelity depletion chain
  results in predictions of $k_\text{eff}$ that agree within 20--30 pcm between
  OpenMC and Serpent. Predicted actinide concentrations were found to agree to a
  fraction of a percent, and most fission product concentrations were found to
  agree within 1\%. The few cases where larger differences were observed can be
  attributed either to differences in how the energy dependence of fission
  product yields is handled or deficiencies in the nuclear data used. OpenMC
  simulations of the PWR and SFR problems using a simplified 228 nuclide
  depletion chain demonstrate that it achieves accuracy close to that of the
  full, high-fidelity depletion chain.
\end{abstract}

\begin{keyword}
  Monte Carlo, particle transport, depletion, transmutation
\end{keyword}

@@

In OpenMC, the burnup matrix is constructed based on data inside of a
\emph{depletion chain} file, which includes fundamental data gathered from ENDF
incident neutron, decay, and fission product yield (FPY) sublibraries. For each
nuclide, this file includes:
\begin{itemize}
  \item What transmutation reactions are possible, their Q values, and their products;
  \item If a nuclide is not stable, what decay modes are possible, their
  branching ratios, and their products; and
  \item If a nuclide is fissionable, the fission products yields at any number
  of incident neutron energies.
\end{itemize}

Depletion chains can be automatically created in the Python API in the following
manner. First, the user supplies a list of ENDF incident neutron, decay, and FPY
sublibrary files. The incident neutron files are used to determine what nuclear
reactions result in transmutation and what the corresponding Q values of the
reactions are. The decay files are used to determine the half-life and possible
decay modes of nuclides that are unstable. The FPY sublibraries are used to
determine what fission products are produced when a fission reaction occurs.
This information is then aggregated and written to an XML file that serves as
the depletion chain file that is used at runtime to construct a burnup matrix.

@@

Some nuclides that have fission reactions in ENDF/B-VII.1, such as $^{243}$Pu,
do not have corresponding FPYs. For these nuclides, a decision has to be made on
what set of FPY should be used when fission occurs. This issue was briefly
studied by \citet{fensin2015pne}, who found that the proportion of fission
reactions that occur in these nuclides is usually small enough that the choice
of FPYs is not of great importance. For the depletion chains used in OpenMC, we
have adopted the method used in Serpent, which is as follows:
\begin{enumerate}
  \item If the nuclide missing FPYs is a metastable state, use FPYs from the
  ground state;
  \item Use FPYs from an isotone if available;
  \item Use FPYs from $^{235}$U.
\end{enumerate}

Using all incident neutron, decay, and FPY sublibrary files from an ENDF-format
library, such as ENDF/B-VIII.0, results in a \emph{full} depletion chain that
typically includes thousands of nuclides. Because the use of a chain with $n$
nuclides will result in an $n\times n$ burnup matrix, there is an incentive to
reduce the number of nuclides in the depletion chain to minimize both the size
of the burnup matrix and the number of nuclides that are tracked during a
transport solve. The depletion chain class in OpenMC's Python API includes a
method for automatically reducing the size of the chain. Starting with a set of
nuclides that must be included, the method will recursively go through all decay
and reaction modes to determine what products may be produced. Using this method
on the ENDF/B-VII.1 library starting from a list of all stable or very long
lived (half-life greater than $10^{15}$ sec) nuclides reduces the total number
of nuclides in the depletion chain from 3,820 to 1,693.

In addition to the full and reduced depletion chains, a simplified depletion
chain based on the VERA depletion benchmark specification~\citep{kim2015casl},
which will henceforth be referred to as the \emph{CASL} depletion chain, was
created and made available for users at \url{https://openmc.org}. This chain
includes 228 nuclides and has adjusted decay branching ratios for some nuclides
to preserve the accuracy of nuclide production rates. \Cref{fig:chains} shows
which nuclides are included in each of the depletion chains. To accurately model
fast reactor problems (which may include Zr and Mo in the fuel), we have
included $^{90}$Zr, $^{92}$Zr, $^{94}$Zr, $^{92}$Mo, and $^{94}$Mo in this
depletion chain even though they were not included in the original
specification.
\begin{figure}
  \centering
  \includegraphics[width=4in]{figures/chains.pdf}
  \caption{Table of nuclides showing which nuclides are included in each of the
    depletion chains based on ENDF/B-VII.1.}
  \label{fig:chains}
\end{figure}

@@

To validate the depletion solver in OpenMC, we have carried out simulations
using the \texttt{openmc.deplete} module as well as a developmental version of
Serpent 2.1.32 on two simple problems. The first problem is a pressurized water
reactor (PWR) pincell and the second problem is a single assembly of a
sodium-cooled fast reactor (SFR). These problems were selected to cover both
thermal and fast neutron spectra. While we had originally attempted to use
Serpent 2.1.31, early comparisons between OpenMC and this version of Serpent
revealed significant differences in the predicted concentration of $^{97}$Mo.
This was found to be due to $^{97\text{m}}$Y having a non-zero independent yield
and a zero cumulative yield in the FPY sublibrary file for $^{235}$U fission in
ENDF/B-VII.1. The zero cumulative yield caused Serpent 2.1.31 to ignore it
altogether, which resulted in significantly less production of $^{97}$Mo. The
developmental version of Serpent 2.1.32 no longer ignores this particular
fission product despite the erroneous cumulative yield.

@@

Six coupled transport-depletion simulations were carried out with OpenMC to test
a variety of combinations of depletion chains and methods for handling FPY
energy dependence:
\begin{enumerate}
  \item Full depletion chain, thermal FPYs
  \item Full depletion chain, interpolated FPYs
  \item Reduced depletion chain, thermal FPYs
  \item CASL depletion chain, thermal FPYs
  %\item CASL depletion chain, FPYs based on cutoff method
  \item CASL depletion chain, interpolated FPYs
\end{enumerate}
Because Serpent doesn't have an option for changing the depletion chain or the
method for handling FPY energy dependence, a single coupled transport-depletion
simulation with Serpent was carried out. For all simulations, 20 inactive
batches, 200 active batches, and \num{4.0e6} particles per batch were used,
which resulted in an uncertainty on the effective multiplication factor of about
3--4 pcm.
